\abstract{%
Artificial-intelligence systems increasingly combine prompts, retrieved context, memory, tools and agent harnesses, yet these components do not by themselves determine whether an available fact applies to the question being answered. A statement may be well supported and still be wrong because it is stale, merely proposed, outside the relevant scope, unknown to the queried observer or true only in another possible world. We call this failure \emph{situation blindness} and define \emph{situation engineering} as the explicit construction, updating, validation and exposure of the query-relative world state that governs evidence applicability. We introduce Situation-Catching Question Answering (SCQA), a reference control architecture that separates situation sensing, state update and response policy, and SituationCatch-Bench, a 4,200-instance diagnostic suite covering temporal change, hidden premises, source conflict, modality, scope, observer knowledge and counterfactual worlds. Under oracle structured states, SCQA satisfies all generated invariants, whereas non-situational rules fail in predictable categories. We interpret this result as a software-level constructive upper bound, not as evidence of end-to-end language-model performance. The main contribution is therefore a falsifiable problem definition, a modular evaluation decomposition and a reproducible research scaffold for measuring where situation sensing, state maintenance, policy selection and generation fail. By locating situation engineering alongside prompt, context, retrieval, memory, tool, harness, evaluation and serving engineering, the framework provides researchers with interfaces, reporting standards and open experiments for developing systems that distinguish relevant evidence from evidence that is actually applicable.
}

\keywords{large language models, situation engineering, evidence applicability, situation awareness, hallucination, trustworthy artificial intelligence, temporal reasoning, question answering, agent systems}
