\label{sec:introduction}

Large language models (LLMs) increasingly answer questions by combining
parametric knowledge with retrieved documents, tools, memory and long
conversational histories. These mechanisms improve evidence coverage, but they
do not guarantee that a system identifies the \emph{situation} in which a fact
is valid. A model may read that a person was appointed and later resigned, yet
still name that person as the current office holder. It may turn a proposal
into a decision, extend a local policy to an entire organization, answer from
information unavailable to the queried observer, or collapse a hypothetical
scenario into the real world. In each case the decisive evidence can already be
present. The failure is not simply missing knowledge; it is incorrect evidence
applicability.

We define \emph{situation blindness} as producing an answer or action that is
supported by at least one available claim but inconsistent with the time, scope,
modality, observer knowledge or possible world required by the query. This
category matters because conventional factuality checks can mark the supporting
claim as true while missing that it is true in the wrong situation. Natural
question-answering resources already show that a material subset of ordinary
questions depends on extra-linguistic time or place, and that updated evidence
does not eliminate the resulting performance loss~\cite{zhang2021situatedqa,
vu2023freshqa,meem2024pat,chen2021timeqa,chen2025chronoqa}.

Retrieval-augmented generation supplies external evidence~\cite{lewis2020rag};
self-reflective and tool-using systems add critique and action
~\cite{asai2024selfrag,yao2023react,schick2023toolformer}; context engineering
organizes the inference-time information payload~\cite{mei2025context}; and
agent harnesses mediate tools, memory, permissions, verification and feedback
~\cite{zhong2026harness,lin2026ahe}. These layers are necessary, but none
requires an explicit answer to a separate question: \emph{which interpretation
of the available evidence governs this interaction now?}

We call the corresponding discipline \emph{situation engineering}: the
systematic design of representations, sensing interfaces, update rules,
validation procedures and response policies that establish the active
query-relative world state before an AI system commits to an answer or action.
For a query $q$ and evidence $E$, a minimal state is
\begin{equation}
S(q,E)=\{\mathcal{A},\mathcal{V},\mathcal{T},\mathcal{P},\mathcal{M},
\mathcal{K},\mathcal{W},\mathcal{U}\},
\end{equation}
where the variables denote actors, events, valid time, scope, modality and
source status, observer-specific knowledge, possible world and unresolved
answer-critical information. A policy then selects
\begin{equation}
\pi(S)\in\{\textsc{Answer},\textsc{Clarify},\textsc{Retrieve},
\textsc{Abstain},\textsc{Act}\}.
\end{equation}
This decomposition separates evidence availability from evidence applicability.

The present paper makes four bounded contributions. First, it defines situation
blindness as a testable failure class and relates it to established work on
situations, temporal reasoning, belief state tracking and knowledge revision.
Second, it specifies situation engineering as a systems contract connecting
prompt, context, retrieval, memory, tool, harness, evaluation and serving
engineering without subsuming them. Third, it introduces SCQA and
SituationCatch-Bench as a transparent oracle-state diagnostic for causal unit
tests and component ablations. Fourth, it provides a staged evaluation and
reporting protocol for future end-to-end studies on natural language, multiple
models, human clarification and deployed agents. We deliberately do not claim
that the current rule-based implementation solves situation sensing in
unrestricted text. Instead, the paper separates the solved control invariants
from the open perception, learning and generalization problems so that future
work can measure progress without conflating them.
