\section*{Related conceptual foundations}

Situation engineering draws on several mature traditions but is not identical
to any one of them. Situation semantics treats meaning relative to partial
situations~\cite{barwise1983situations}. Situation and event calculi formalize
how actions change fluents over time~\cite{mccarthy1969problems,
kowalski1986event,reiter2001knowledge}. Temporal reasoning provides interval
relations and validity semantics~\cite{allen1983interval}. Belief revision and
truth-maintenance research asks how an agent should update inconsistent
commitments~\cite{alchourron1985logic,doyle1979truth}. POMDPs and dialogue-state
tracking maintain uncertainty over latent states for sequential decisions
~\cite{kaelbling1998pomdp,henderson2014second}. Temporal knowledge graphs and
bitemporal databases record changing facts and distinguish when a fact is valid
from when it is stored~\cite{snodgrass1992temporal}.

The proposed contribution is a systems-level applicability contract for
generative and agentic AI. It requires these representations to connect to
natural-language evidence, provenance, unresolved alternatives and a response
policy that can answer, retrieve, clarify, abstain or act. Thus the novelty is
not a new universal logic. It is the explicit decomposition of modern AI
reliability into (i) sensing a situation from heterogeneous evidence, (ii)
maintaining a provenance-linked state, (iii) validating applicability and
(iv) propagating unresolved state into behaviour. This framing allows existing
logical, probabilistic and neural methods to be compared through a common set of
interfaces and failure measurements.

\begin{table}[t]
\caption{Typical coverage of neighbouring state-based research traditions.
Entries summarize common formulations rather than claiming that every method in
an area has the same capabilities.}
\label{tab:foundations}
\centering\scriptsize
\setlength{\tabcolsep}{3pt}%
\begin{tabularx}{\linewidth}{@{}p{2.1cm}*{7}{>{\hspace{0pt}\centering\arraybackslash}X}@{}}
\toprule
Area & Explicit state & Time & Observer & Worlds & Uncertainty & Answer/\allowbreak action gate & Provenance \\
\midrule
Situation/\allowbreak event calculus & $\checkmark$ & $\checkmark$ & limited & partial & usually no & action-centred & limited \\
Belief revision / truth maintenance & $\checkmark$ & partial & partial & partial & partial & no & limited \\
Dialogue-state tracking / POMDPs & $\checkmark$ & sequential & $\checkmark$ & partial & $\checkmark$ & $\checkmark$ & uncommon \\
Temporal KG / bitemporal data & $\checkmark$ & $\checkmark$ & limited & uncommon & limited & no & $\checkmark$ \\
RAG / tool agents & often implicit & limited & uncommon & uncommon & model-derived & partial & partial \\
Proposed situation interface & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & required & $\checkmark$ & $\checkmark$ \\
\bottomrule
\end{tabularx}
\end{table}

Dynamic epistemic logic and common-ground modelling formalize belief-changing
information and mutually recognized conversational commitments
~\cite{vanbenthem1996dynamic,stalnaker2002common}. Theory-of-mind models, world
models and active perception further address attributed beliefs, latent
environment state and information acquisition. These traditions reinforce rather
than eliminate the need for careful attribution: situation engineering does not
claim their theoretical results as new. Its proposed novelty is the systems
contract that exposes their state variables to provenance-aware
answer/retrieve/clarify/abstain gates in contemporary generative systems.
