\section*{From prompt engineering to situation engineering}

The reliability stack surrounding foundation models has expanded in response to
successive failure modes. Prompt engineering shapes the immediate instruction:
what task to perform, which role to assume and how to express the output.
Context engineering expands the unit of design to the information payload,
including retrieval, memory, examples, tool observations and context-window
management~\cite{mei2025context}. Harness engineering expands it again to the
runtime substrate that mediates actions, permissions, verification, persistence
and feedback~\cite{zhong2026harness,lin2026ahe}. These layers are complementary,
but none necessarily constructs an explicit answer to a different question:
\emph{which interpretation of the available evidence governs the present
interaction?}

We use \emph{situation engineering} for the systematic design of representations,
update rules, sensing processes and decision policies that establish the active
world state before an AI system commits to an answer or action. The term is not
intended to replace context, memory, retrieval, tools, evaluation or harness
engineering. It identifies the semantic control plane between information
availability and justified behaviour. Context engineering can place an
appointment notice and a later resignation in the same window. Retrieval
engineering can find both. Memory engineering can retain both across sessions.
Tool engineering can expose an authoritative registry. Harness engineering can
guarantee that the records are fetched, logged and checked. Evaluation
engineering can detect that the final answer is wrong. Situation engineering is
the layer that determines, before commitment, that the resignation supersedes
the appointment for a question about the current office holder.

\subsection*{A layered taxonomy of AI systems engineering}

The emerging vocabulary of ``engineering'' around generative AI mixes design
objects that operate at different levels. Treating every term as a peer produces
both overlap and conceptual inflation. We therefore organize the stack into
three interacting planes (Fig.~\ref{fig:engineering-planes}). The
\emph{semantic and cognitive plane} determines instructions, evidence, memory
and the active situation. The \emph{action and runtime plane} determines which
capabilities can be invoked and how multi-step execution is governed. The
\emph{measurement and resource plane} determines how behaviour is evaluated and
how inference is delivered under latency, throughput, cost and energy budgets.
This grouping is functional rather than organizational: one production team or
one software framework may implement several planes.

\begin{figure}[t]
\centering
\begin{tikzpicture}[font=\small, node distance=4mm and 5mm,
box/.style={draw,rounded corners,align=center,minimum height=8mm,text width=2.75cm,fill=white},
plane/.style={draw,rounded corners,inner sep=5mm}]
\node[box] (prompt) {Prompt\\engineering};
\node[box,right=of prompt] (context) {Context and retrieval\\engineering};
\node[box,right=of context] (memory) {Memory\\engineering};
\node[box,right=of memory,draw=black,very thick] (situation) {Situation\\engineering};
\node[plane,fit=(prompt)(context)(memory)(situation),label=above:{Semantic and cognitive plane}] (p1) {};

\node[box,below=12mm of prompt] (tool) {Tool / function-calling\\engineering};
\node[box,right=of tool] (protocol) {Protocol and interface\\engineering};
\node[box,right=of protocol] (harness) {Harness and orchestration\\engineering};
\node[box,right=of harness] (safety) {Guardrail and security\\engineering};
\node[plane,fit=(tool)(protocol)(harness)(safety),label=above:{Action and runtime plane}] (p2) {};

\node[box,below=12mm of tool] (eval) {Evaluation\\engineering};
\node[box,right=of eval] (obs) {Observability and provenance\\engineering};
\node[box,right=of obs] (serve) {Inference / serving\\engineering};
\node[box,right=of serve] (ops) {Reliability and cost\\engineering};
\node[plane,fit=(eval)(obs)(serve)(ops),label=above:{Measurement and resource plane}] (p3) {};

\draw[-{Latex}] (prompt) -- (context);
\draw[-{Latex}] (context) -- (memory);
\draw[-{Latex}] (memory) -- (situation);
\draw[-{Latex}] (tool) -- (harness);
\draw[-{Latex}] (protocol) -- (harness);
\draw[-{Latex}] (harness) -- (safety);
\draw[-{Latex}] (eval) -- (obs);
\draw[-{Latex}] (serve) -- (ops);
\draw[-{Latex},dashed] (context) -- (tool);
\draw[-{Latex},dashed] (memory) -- (harness);
\draw[-{Latex},very thick] (situation) -- (harness);
\draw[-{Latex},very thick] (harness) -- (situation);
\draw[-{Latex},dashed] (situation) -- (eval);
\draw[-{Latex},dashed] (obs) -- (situation);
\end{tikzpicture}
\caption{A functional taxonomy of engineering disciplines surrounding modern AI systems. Solid horizontal arrows indicate common within-plane dependencies; dashed arrows indicate cross-plane feedback. Situation engineering is distinguished by its responsibility for evidence applicability and active-world-state control, not by ownership of a particular software component.}
\label{fig:engineering-planes}
\end{figure}

\begin{table}[t]
\centering
\caption{Complementary engineering disciplines for reliable AI. Boundaries are
analytical, and substantial implementation overlap is expected.}
\label{tab:engineering-stack}
\begin{tabularx}{\linewidth}{p{2.8cm}p{3.0cm}X p{3.0cm}}
\toprule
Discipline & Primary design object & Reliability question & Characteristic failure if absent \\
\midrule
Prompt engineering & Instructions, roles and output constraints & Did the model understand what it was asked to do? & Wrong task, format or reasoning procedure \\
Context engineering & Evidence payload and context-window allocation & Did the model receive the information needed to do it? & Missing, irrelevant or poorly organized evidence \\
Retrieval engineering & Query formulation, indexes, ranking and freshness & Was the best available evidence found efficiently? & Relevant evidence omitted or stale evidence over-ranked \\
Memory engineering & Write, consolidation, forgetting and recall policies & What should persist across turns and sessions? & Lost commitments, repeated mistakes or stale memories \\
Situation engineering & Time-indexed, scoped, observer-relative world state & Which available facts apply here and now, and what remains unresolved? & Confident use of expired, out-of-scope, hypothetical or unknown facts \\
Tool / function-calling engineering & Capability schemas, descriptions, preconditions and error semantics & Can the model select and invoke the right capability correctly? & Wrong tool, malformed arguments or silent tool failure \\
Protocol / interface engineering & Interoperability, lifecycle, authorization and capability negotiation & Can components exchange context and actions safely and consistently? & Integration mismatch, privilege confusion or ambiguous contracts \\
Harness engineering & Orchestration, permissions, retries, checkpoints and recovery & Can the system execute a multi-step workflow safely? & Unchecked actions, brittle loops or unrecoverable execution \\
Evaluation engineering & Tasks, rubrics, judges, counterfactuals and regression suites & Can reliability claims be measured and falsified? & Hidden regressions, metric gaming or unsupported claims \\
Observability / provenance engineering & Traces, state snapshots, lineage and audit records & Can an operator reconstruct why the output occurred? & Un-debuggable failures and unverifiable decisions \\
Inference / serving engineering & Scheduling, batching, caching, placement and numerical execution & Can the system meet latency, throughput, cost and energy constraints? & Timeouts, overload, excessive cost or quality loss under load \\
Guardrail / security engineering & Policy enforcement, isolation, authorization and abuse resistance & Are actions and data flows constrained to acceptable risk? & Unsafe action, leakage, injection or privilege escalation \\
\bottomrule
\end{tabularx}
\end{table}

Several relationships are especially important. Retrieval engineering is a
specialized part of context acquisition, but also has its own ranking,
indexing, freshness and systems questions. Memory engineering stores and
recovers information, whereas situation engineering decides whether a recovered
memory is still valid, is superseded, belongs to the current user or world, or
should remain only a competing hypothesis. Tool engineering specifies
capabilities and invocation contracts; harness engineering composes those
capabilities into guarded workflows. Protocols such as the Model Context
Protocol standardize how prompts, resources and tools are exposed, but a
protocol does not by itself determine the semantic applicability of the data it
transports~\cite{mcp2025}. Evaluation engineering is not merely a final score:
it is the sensing system for the engineering lifecycle, translating desired
behaviour into tests, traces and regression signals~\cite{anthropic2026evals,openai2026evals}.
Inference and serving engineering are orthogonal but consequential because
resource pressure can alter truncation, retrieval depth, model choice and
verification budgets, thereby indirectly changing situation quality~\cite{li2024serving}.

\subsection*{Situation engineering as an evidence-applicability discipline}

A situation is not a longer prompt and not merely a collection of metadata. It
is a constrained interpretation of evidence relative to a query. Let $E$ be the
available evidence and $q$ a query. Context selection produces a subset or
representation $C(q,E)$. Memory contributes retained observations $M_t$.
Tools produce new observations $O_t$, and a harness supplies an execution trace
$H_t$. Situation construction produces an active state
$S_t(q,C,M_t,O_t,H_t)$ containing the variables that decide whether each claim
is applicable. Generation should therefore be conditioned not only on evidence
relevance but on an applicability relation
\begin{equation}
\mathrm{Apply}(e,q,S_t)\in\{\mathrm{valid},\mathrm{invalid},\mathrm{unresolved}\}.
\end{equation}
A reliable system must preserve the third value. Converting unresolved
applicability into a plausible binary judgement is one mechanism by which
fluent systems hallucinate.

We distinguish six operations. \emph{Situation sensing} extracts events,
actors, times, scopes, source status, perspectives and world identifiers from
language, databases, tools or sensors. \emph{Situation assembly} links those
observations into candidate states. \emph{Situation updating} applies event and
belief transitions as new evidence arrives. \emph{Situation validation} checks
consistency, provenance, freshness and unresolved alternatives. \emph{Situation
policy} selects whether to answer, retrieve, clarify, act or abstain. Finally,
\emph{situation observability} records which state and evidence justified the
decision. SCQA operationalizes a minimal version of these operations for
question answering.

\subsection*{Why model scale, tools and longer context are insufficient}

Increasing model capacity may improve event extraction and implicit reasoning,
but it does not create a contract that the active state must be represented,
updated and inspected. Longer windows can contain more mutually incompatible
states: old and new policies, proposals and final decisions, public facts and
private beliefs, or real and hypothetical scenarios. More tools can expand both
competence and the consequences of applying an incorrect state. More memory can
preserve both useful commitments and obsolete assumptions. The reliability
objective is therefore not to maximize information volume or capability count,
but to minimize answer-critical unresolved applicability subject to retaining
the evidence and actions needed for the task.

A compact systems formulation is
\begin{equation}
\begin{split}
\text{trustworthy outcome}={}&\text{task alignment}+\text{evidence coverage}
+\text{state continuity}\\
&+\text{capability correctness}+\text{runtime control}
+\text{validated situation state}\\
&+\text{measurement and auditability}+\text{resource feasibility}.
\end{split}
\end{equation}
These terms map respectively to prompt, context/retrieval, memory, tool,
harness/security, situation, evaluation/observability and serving engineering.
The expression is a design decomposition rather than an additive statistical
model: failure in any term can dominate the outcome.

\subsection*{Situation engineering as the coordinator of state-bearing components}

Situation engineering has particularly close interfaces with four neighbouring
disciplines. First, retrieval should be driven by missing situation variables,
not only semantic similarity. A system uncertain about whether a policy is
active should retrieve enactment, effective-date and repeal evidence rather than
more documents that merely discuss the policy. Second, memory should be event-
and provenance-aware. It should store when a fact was learned, when it was valid,
which agent observed it and what later event superseded it~\cite{du2026memory}.
Third, tools should declare state effects and preconditions, analogous to typed
transactions: a booking tool should report whether a reservation is held,
confirmed, expired or cancelled, not merely return free text. Fourth, harnesses
should enforce situation invariants before consequential action, for example
blocking payment when identity, amount, authorization scope or transaction world
remains unresolved.

This coordinator role does not make situation engineering a universal container
for all other engineering. It owns the semantics of the active state and the
policy consequences of uncertainty in that state. The retriever still owns
ranking quality; the memory subsystem still owns retention and forgetting; the
tool layer still owns safe invocation; the harness still owns workflow integrity;
the evaluator still owns valid measurement; and the serving layer still owns
resource efficiency. A useful implementation contract is therefore a typed
interface rather than organizational consolidation:
\begin{equation}
(C_t,M_t,O_t,H_t)\rightarrow (\mathcal{S}_t,U_t,P_t),
\end{equation}
where $\mathcal{S}_t$ is a set of candidate situations, $U_t$ their uncertainty
and unresolved variables, and $P_t$ the permitted response or action policy.

\subsection*{A maturity model for situation-engineered AI}

We propose five levels that turn the concept into a falsifiable systems agenda.
At SE0, the system has no explicit situation representation and relies on
language-model inference alone. At SE1, it tags isolated variables such as time
or location. At SE2, it maintains an explicit state with provenance and update
rules. At SE3, it detects missing variables, competing hypotheses and
contradictions, and can clarify or abstain. At SE4, it learns and audits
situation states across modalities, agents and long-running environments while
quantifying uncertainty and distribution shift. SituationCatch-Bench primarily
tests the transition from SE0/SE1 to SE2/SE3 under controlled conditions.

This maturity model prevents a common category error. A system that prints a
JSON field called ``situation'' is not necessarily situation-aware. Evidence is
required that the representation causally changes behaviour in the predicted
cases, that ablations recover matching failure signatures, and that uncertainty
in the state propagates to uncertainty in the answer. The experiments below
provide this causal diagnostic for seven variables, while leaving natural-text
sensing and SE4 operation as open problems.

\subsection*{Research agenda for epistemically regulated AI}

A useful reliability target is not zero error but appropriate epistemic regulation. People ask for
clarification, track who knows what, revise beliefs, recognize hypotheticals and
withhold judgement when the situation is underspecified. A useful computational
target is therefore \emph{human-like epistemic regulation}: the ability to
identify which state is active, notice when it cannot be determined and choose
a proportionate response. Progress requires connected research programmes
across the engineering planes.

Neural situation sensing must extract implicit and distributed state variables
from natural text, speech, images and operational traces. Dynamic state models
must represent valid time, transaction time, event supersession, causal
dependencies and observer-relative beliefs without erasing uncertainty. Active
information acquisition must select the retrieval, memory recall, tool call or
clarification that most reduces answer-critical uncertainty. Situation-
conditioned calibration must estimate risk from state completeness, conflict and
provenance in addition to model logits. Evaluation engineering must separately
score sensing, updating, policy and final response under adversarially stale or
manipulated evidence. Serving research must study how latency and cost budgets
change the achievable depth of verification and state maintenance.

These programmes suggest that situation engineering can serve as a common
semantic interface among otherwise separate communities studying RAG, temporal
reasoning, knowledge graphs, agent memory, tools, protocols, agent harnesses,
uncertainty, human--AI interaction, serving systems and AI safety. An event-
memory system can expose a situation-state API; a retriever can optimize for
missing variables; a tool can publish preconditions and state transitions; a
harness can prevent action until state invariants pass; an evaluator can probe
state-specific counterfactuals; and a generator can cite the state and evidence
on which each claim depends. The practical goal is not a new label for existing
components, but an auditable contract: no consequential answer or action should
be produced without an explicit account of the situation in which it is valid.
