\section*{Discussion}

The core insight is that relevance is not applicability. Retrieval can return a
well-supported statement while a system still applies it to the wrong time,
scope, epistemic state or possible world. Situation engineering turns this
failure from an informal complaint about ``missing context'' into a set of
state variables, update rules and behavioural consequences that can be tested
separately.

\subsection*{A complementary layer, not a replacement vocabulary}

Prompt, context, retrieval, memory, tool, harness, evaluation, security and
serving engineering retain distinct responsibilities. Prompts specify the task;
context and retrieval supply evidence; memory preserves observations; tools
provide capabilities; harnesses govern workflow and permissions; evaluations
measure behaviour; serving systems manage resource constraints. Situation
engineering owns the query-relative semantics of the active state and the
propagation of unresolved applicability into policy. A useful contract is
\begin{equation}
(C_t,M_t,O_t,H_t)\rightarrow(\mathcal{S}_t,U_t,P_t),
\end{equation}
where context, memory, tool observations and harness traces produce candidate
situations $\mathcal{S}_t$, unresolved variables $U_t$ and the permitted policy
$P_t$.

This contract prevents conceptual inflation. Retrieval quality should not be
renamed situation engineering, and a JSON field named ``situation'' is not
evidence of situation awareness. A system must show that state representation
causally changes behaviour in predicted cases, that state uncertainty affects
answering or action, and that failures can be assigned to sensing, update,
validation, policy or generation.

\subsection*{Implications for hallucination and agent evaluation}

A generated claim can be textually entailed by a document yet wrong for the
query. Evaluation should therefore record both support and applicability. In
free-form generation, claims should be linked to the situation state and
provenance on which they depend. In agents, tools should expose preconditions,
state transitions and failure semantics rather than only free-text outputs;
harnesses should block consequential actions while answer-critical state is
unresolved. Evaluation engineering becomes the sensor layer that checks each
transition and not merely the final answer.

The response policy is also part of correctness. Clarification and abstention
are not failures when the situation is under-specified. Conversely, excessive
clarification is costly and can make systems unusable. Future evaluations should
therefore report answer accuracy, clarification utility, unnecessary-question
rate, abstention precision, selective risk, coverage, latency, token use and
monetary cost. These measures make explicit the trade-off between semantic
verification and serving efficiency.

\subsection*{What is established, supported and proposed}

This study establishes a formal error class, an interoperable state schema, a
reference implementation and matched oracle-state ablations. Independent
literature supports the premise that time- and place-dependent questions are
common enough to matter. The study does not establish that current frontier
models exhibit every proposed category at a particular frequency, nor that SCQA
improves them under natural language. Those are hypotheses enabled by the
released protocol, not completed results.

This distinction changes the scientific role of the paper. The contribution is
not a claim of a production-ready solution. It is a research scaffold that
allows subsequent work to ask more precise questions: Is the dominant error in
state sensing or state update? Does explicit state outperform a matched
structured prompt? Does a temporal graph add value after controlling for token
budget? When should a system ask rather than retrieve? How robust is a state to
adversarially stale evidence? Which costs arise from maintaining provenance and
multiple candidate worlds?

\subsection*{A staged research programme}

We propose six stages for cumulative evidence. Stage 1 uses formal and synthetic
unit tests to verify semantics. Stage 2 evaluates natural-language situation
sensing with slot, event, temporal, scope, modality, observer and world-state
metrics. Stage 3 performs matched interventions on several model families while
holding evidence and computation budgets constant. Stage 4 tests compositional
and distributional generalization: unseen templates, longer event chains,
nested scope, conflicting authoritative sources, multiple observers,
multilingual data and multimodal observations. Stage 5 adds human annotation and
interaction to measure clarification and abstention. Stage 6 evaluates
long-running agents, state drift, security attacks, latency, cost and independent
auditing.

The SE0--SE4 maturity model and longer research agenda are moved to the
Supplementary Information. The present implementation reaches explicit
state-and-policy control only for seven supplied variables.

\subsection*{Limitations and risks}

The diagnostic data are generated from a schema shared with the solver, so
benchmark-solver circularity is intentional and limits external validity. The
item count overstates statistical independence because examples share template
families. The baselines are software controls, not state-of-the-art LLM
systems. Confidence values are rule-derived and do not constitute neural
calibration. No human participants, multilingual evaluation, multimodal sensing,
frontier-model calls or deployed-agent study were completed. These limitations
are not resolved by additional prose and require new experiments.

Explicit state can also create new attack surfaces. An adversary may manipulate
timestamps, source authority, identity, scope or world labels. A centralized
state representation may expose sensitive user information or create false
confidence in a flawed schema. Situation-engineered systems therefore require
provenance, access control, privacy-preserving memory, conflict visibility and
independent evaluation. ``Explicit'' must not be confused with ``correct''.

\subsection*{Conclusion}

Situation engineering provides a bounded and testable way to connect several
communities working on reliable AI. Its durable contribution will depend not on
the adoption of a new label, but on whether researchers use the interfaces to
produce stronger measurements, interoperable state representations and systems
that know when available evidence does and does not apply. The present artefact
is intended as the first rung of that evidence ladder: a transparent formal and
software scaffold on which natural-language, multi-model and deployment studies
can be built.
