\begin{thebibliography}{40}
\bibitem{lewis2020rag} Lewis, P. et al. Retrieval-augmented generation for knowledge-intensive NLP tasks. \emph{Adv. Neural Inf. Process. Syst.} \textbf{33}, 9459--9474 (2020).
\bibitem{asai2024selfrag} Asai, A., Wu, Z., Wang, Y., Sil, A. \& Hajishirzi, H. Self-RAG: learning to retrieve, generate, and critique through self-reflection. In \emph{Proc. International Conference on Learning Representations} (2024).
\bibitem{yao2023react} Yao, S. et al. ReAct: synergizing reasoning and acting in language models. In \emph{Proc. International Conference on Learning Representations} (2023).
\bibitem{schick2023toolformer} Schick, T. et al. Toolformer: language models can teach themselves to use tools. \emph{Adv. Neural Inf. Process. Syst.} \textbf{36} (2023).
\bibitem{min2020ambigqa} Min, S., Michael, J., Hajishirzi, H. \& Zettlemoyer, L. AmbigQA: answering ambiguous open-domain questions. In \emph{Proc. EMNLP}, 5783--5797 (2020).
\bibitem{zhang2021situatedqa} Zhang, M. J. Q. \& Choi, E. SituatedQA: incorporating extra-linguistic contexts into QA. In \emph{Proc. EMNLP}, 7371--7387 (2021).
\bibitem{kamath2020selectiveqa} Kamath, A., Jia, R. \& Liang, P. Selective question answering under domain shift. In \emph{Proc. ACL}, 5684--5696 (2020).
\bibitem{lin2022truthfulqa} Lin, S., Hilton, J. \& Evans, O. TruthfulQA: measuring how models mimic human falsehoods. In \emph{Proc. ACL}, 3214--3252 (2022).
\bibitem{ji2023hallucination} Ji, Z. et al. Survey of hallucination in natural language generation. \emph{ACM Comput. Surv.} \textbf{55}, 1--38 (2023).
\bibitem{guo2017calibration} Guo, C., Pleiss, G., Sun, Y. \& Weinberger, K. Q. On calibration of modern neural networks. In \emph{Proc. ICML}, 1321--1330 (2017).
\bibitem{manakul2023selfcheckgpt} Manakul, P., Liusie, A. \& Gales, M. J. F. SelfCheckGPT: zero-resource black-box hallucination detection for generative large language models. In \emph{Proc. EMNLP}, 9004--9017 (2023).
\bibitem{thorne2018fever} Thorne, J., Vlachos, A., Christodoulopoulos, C. \& Mittal, A. FEVER: a large-scale dataset for fact extraction and verification. In \emph{Proc. NAACL-HLT}, 809--819 (2018).
\bibitem{allen1983interval} Allen, J. F. Maintaining knowledge about temporal intervals. \emph{Commun. ACM} \textbf{26}, 832--843 (1983).
\bibitem{pearl1988probabilistic} Pearl, J. \emph{Probabilistic Reasoning in Intelligent Systems} (Morgan Kaufmann, 1988).
\bibitem{wei2022cot} Wei, J. et al. Chain-of-thought prompting elicits reasoning in large language models. \emph{Adv. Neural Inf. Process. Syst.} \textbf{35}, 24824--24837 (2022).
\bibitem{mei2025context} Mei, L. et al. A survey of context engineering for large language models. \emph{Preprint at} \url{https://arxiv.org/abs/2507.13334} (2025).
\bibitem{zhong2026harness} Zhong, H. \& Zhu, S. AI harness engineering: a runtime substrate for foundation-model software agents. \emph{Preprint at} \url{https://arxiv.org/abs/2605.13357} (2026).
\bibitem{lin2026ahe} Lin, J. et al. Agentic harness engineering: observability-driven automatic evolution of coding-agent harnesses. \emph{Preprint at} \url{https://arxiv.org/abs/2604.25850} (2026).
\bibitem{barwise1983situations} Barwise, J. \& Perry, J. \emph{Situations and Attitudes} (MIT Press, 1983).
\bibitem{madaan2023selfrefine} Madaan, A. et al. Self-refine: iterative refinement with self-feedback. \emph{Adv. Neural Inf. Process. Syst.} \textbf{36} (2023).
\bibitem{vu2023freshqa} Vu, T. et al. FreshLLMs: refreshing large language models with search engine augmentation. \emph{Preprint at} \url{https://arxiv.org/abs/2310.03214} (2023).
\bibitem{meem2024pat} Meem, J. et al. PAT-Questions: a self-updating benchmark for present-anchored temporal question answering. In \emph{Findings of the Association for Computational Linguistics: ACL} (2024).
\bibitem{chen2021timeqa} Chen, W., Wang, X., Wang, W. Y. \& Chen, W. A dataset for answering time-sensitive questions. In \emph{Proc. NeurIPS Datasets and Benchmarks} (2021).
\bibitem{chen2025chronoqa} Chen, Z. et al. A question answering dataset for temporal-sensitive retrieval-augmented generation. \emph{Sci. Data} (2025).
\bibitem{kim2025tdbench} Kim, S., Wang, J., Xie, X. \& Whang, S. E. Harnessing temporal databases for systematic evaluation of factual time-sensitive question answering in large language models. \emph{Preprint at} \url{https://arxiv.org/abs/2508.02045} (2025).
\bibitem{mcp2025} Model Context Protocol contributors. Model Context Protocol specification, revision 2025-11-25. \url{https://modelcontextprotocol.io/specification/2025-11-25} (2025).
\bibitem{anthropic2026evals} Anthropic. Demystifying evals for AI agents. \url{https://www.anthropic.com/engineering/demystifying-evals-for-ai-agents} (2026).
\bibitem{openai2026evals} OpenAI. A shared playbook for trustworthy third-party evaluations. \url{https://openai.com/index/trustworthy-third-party-evaluations-foundations/} (2026).
\bibitem{li2024serving} Li, B., Jiang, Y., Gadepally, V. \& Tiwari, D. LLM inference serving: survey of recent advances and opportunities. \emph{Preprint at} \url{https://arxiv.org/abs/2407.12391} (2024).
\bibitem{du2026memory} Du, P. Memory for autonomous LLM agents: mechanisms, evaluation, and emerging frontiers. \emph{Preprint at} \url{https://arxiv.org/abs/2603.07670} (2026).

\bibitem{mccarthy1969problems} McCarthy, J. \& Hayes, P. J. Some philosophical problems from the standpoint of artificial intelligence. In \emph{Machine Intelligence 4}, 463--502 (Edinburgh Univ. Press, 1969).
\bibitem{kowalski1986event} Kowalski, R. \& Sergot, M. A logic-based calculus of events. \emph{New Gener. Comput.} \textbf{4}, 67--95 (1986).
\bibitem{reiter2001knowledge} Reiter, R. \emph{Knowledge in Action: Logical Foundations for Specifying and Implementing Dynamical Systems} (MIT Press, 2001).
\bibitem{alchourron1985logic} Alchourr\'{o}n, C. E., G\"ardenfors, P. \& Makinson, D. On the logic of theory change: partial meet contraction and revision functions. \emph{J. Symb. Logic} \textbf{50}, 510--530 (1985).
\bibitem{doyle1979truth} Doyle, J. A truth maintenance system. \emph{Artif. Intell.} \textbf{12}, 231--272 (1979).
\bibitem{kaelbling1998pomdp} Kaelbling, L. P., Littman, M. L. \& Cassandra, A. R. Planning and acting in partially observable stochastic domains. \emph{Artif. Intell.} \textbf{101}, 99--134 (1998).
\bibitem{henderson2014second} Henderson, M., Thomson, B. \& Williams, J. D. The second dialog state tracking challenge. In \emph{Proc. SIGDIAL}, 263--272 (2014).
\bibitem{snodgrass1992temporal} Snodgrass, R. T. Temporal databases. In \emph{Theories and Methods of Spatio-Temporal Reasoning in Geographic Space}, 22--64 (Springer, 1992).
\end{thebibliography}
