%% ============================================================
%%  Springer Nature / Nature-family journal template
%%  Class file required: sn-jnl.cls
%%  Download: https://www.springernature.com/gp/authors/campaigns/latex-author-support
%%  Target journal: Nature Machine Intelligence  (primary submission target)
%%
%%  This is the single entry file for the paper. Build: ./run.sh
%%  (pdflatex → bibtex → pdflatex × 2  →  main.pdf)
%%  NMI reframing toggle \ifNMIframing (default on); see README.md §4 / §8.
%% ============================================================

\documentclass[
    %% --- journal style options (pick one) ---
    % iicol,          % two-column (IEEE-like preview; final layout set by editors)
    Numbered,         % numbered references  ← Nature style
]{sn-jnl}

%% ---- Springer Nature bibliography style -----------------
%% sn-nature          : Nature-family superscript numeric
%% sn-mathphys-num    : numbered, brackets
%% sn-aps             : APS style
\usepackage[numbers,sort&compress]{natbib}

%% ---- Standard packages ----------------------------------
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue,
bookmarks=true, bookmarksopen=true,
pdftitle={Situation engineering for evidence-applicable artificial intelligence},
pdfauthor={Geunsik Lim},
pdfsubject={Situation engineering, evidence applicability and reliable AI},
pdfkeywords={situation engineering, evidence applicability, reliable AI, temporal reasoning, hallucination}}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{makecell}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{pifont}
\usepackage{soul}
\usepackage{listings}
\usepackage{float}
\usepackage{placeins}
\usepackage{ifthen}
\usepackage{diagbox}

%% ---- Experiment-reporting helpers (kept from original) ---
\newcommand{\ntrials}{5}
\newcommand{\ci}{95\%~CI}
\newcommand{\hwsetup}{2~$\times$~NVIDIA~A100~(80\,GB), 512\,GB DRAM, NVMe Gen4}
\newcommand{\swsetup}{PyTorch~2.4, CUDA~12.2, vLLM~\cite{kwon2023pagedattention}~0.5.3}
\newcommand{\capinfo}{\emph{Means across $n=\ntrials$ runs (\ci). HW: \hwsetup; SW: \swsetup.}}

%% ---- ORCID icon (optional) ------------------------------
\usepackage{scalerel}
\usepackage{tikz}
\usetikzlibrary{svg.path,positioning,arrows.meta,fit}
\definecolor{orcidlogocol}{HTML}{A6CE39}
\tikzset{
  orcidlogo/.pic={
    \fill[orcidlogocol] svg{M256,128c0,70.7-57.3,128-128,128C57.3,256,0,198.7,0,128C0,57.3,57.3,0,128,0C198.7,0,256,57.3,256,128z};
    \fill[white] svg{M86.3,186.2H70.9V79.1h15.4v48.4V186.2z}
                 svg{M108.9,79.1h41.6c39.6,0,57,28.3,57,53.6c0,27.5-21.5,53.6-56.8,53.6h-41.8V79.1z M124.3,172.4h24.5c34.9,0,42.9-26.5,42.9-39.7c0-21.5-13.7-39.7-43.7-39.7h-23.7V172.4z}
                 svg{M88.7,56.8c0,5.5-4.5,10.1-10.1,10.1c-5.6,0-10.1-4.6-10.1-10.1c0-5.6,4.5-10.1,10.1-10.1C84.2,46.7,88.7,51.3,88.7,56.8z};
  }
}
\newcommand\orcidicon[1]{\href{https://orcid.org/#1}{\mbox{\scalerel*{
\begin{tikzpicture}[yscale=-1,transform shape]
\pic{orcidlogo};
\end{tikzpicture}
}{|}}}}

%% ---- Anonymity switch (1 = named, 0 = anonymous) --------
\newcommand{\anonymous}{0}

%% =============================================================
%% NMI framing toggle
%% -------------------------------------------------------------
%% 1차 투고 대상 = Nature Machine Intelligence.
%% \NMIframingtrue  → NMI용 재프레이밍(초록·서론) 사용 (기본값).
%% \NMIframingfalse → 기존 NCS/Nature 서사로 복귀.
%% 근거·전략: README.md §8·§9 참조.
\newif\ifNMIframing
\NMIframingtrue
%% =============================================================

%% =============================================================
%% Reference URL toggle
%% -------------------------------------------------------------
%% 리뷰(열람)용 사본에는 각 참고문헌의 URL을 출력한다 (기본값).
%% \RefURLstrue  → 참고문헌마다 URL 출력 (사전 열람용).
%% \RefURLsfalse → 저널 제출용: URL 출력 안 함.
\newif\ifRefURLs
\RefURLstrue
\newcommand{\refurl}[1]{\ifRefURLs\ \url{#1}\fi}
%% =============================================================

%% =============================================================
\begin{document}
%% =============================================================

%% ---- Title -------------------------------------------------
\input{./section/001_title.tex}

%% ---- Authors & Affiliations --------------------------------
\input{./section/005_author_nature.tex}

%% ---- Abstract & Keywords -----------------------------------
\ifNMIframing
  \input{./section/006_abstract_nmi.tex}      % NMI framing (Nature syntax)
\else
  \input{./section/006_abstract_nature.tex}   % original NCS/Nature abstract
\fi

%% sn-jnl emits the abstract inside \maketitle, so it must come after
%% the \abstract{...} definition above.
\maketitle

%% ---- Main text (NCS structure: Intro → Results → Discussion) ---
\ifNMIframing
  \pdfbookmark[1]{Introduction}{bookmark-introduction}
  \input{section/010_introduction_nmi}         % NMI-reframed introduction
\else
  \input{section/010_introduction}             % original introduction
\fi
\pdfbookmark[1]{Related Work}{bookmark-related}
\input{section/012_related_work}
\pdfbookmark[1]{Situation Engineering Framework}{bookmark-framework}
\input{section/015_situation_engineering}
\pdfbookmark[1]{Results}{bookmark-results}
\input{section/025_results_ncs}
\pdfbookmark[1]{Discussion}{bookmark-discussion}
\input{section/060_discussion}

%% ---- Methods (Nature: starred section, not counted in word limit) ----
\pdfbookmark[1]{Methods}{bookmark-methods}
\input{section/070_methods}

%% ---- Data availability (Nature Portfolio: standalone statement) ----
\section*{Data availability}
SituationCatch-Bench v0.1 is included in the public source repository at
\url{https://github.com/leemgs/sage} and in the submission package as
\texttt{data/situationcatch\_bench.jsonl}, together with its deterministic
generator, datasheet and item-level outputs. This package is sufficient to
reproduce the controlled results. The exact reviewed version is identified by
the Git commit attached to the submission. An immutable archival release, an
institutionally approved licence and DOI remain required before acceptance; no
placeholder DOI is represented as an existing deposit.

%% ---- Code availability (Nature Portfolio: standalone statement) ----
\section*{Code availability}
The complete executable implementation is included under \texttt{code/}.
Running \texttt{PYTHONPATH=code python code/run\_experiments.py}
regenerates gold-, predicted- and corrupted-state outputs without network
access or proprietary dependencies. The multi-model adapter records raw
responses and errors and never creates synthetic API results. The annotation
program requires three complete independent packets before reporting agreement.
Source and item-level outputs are also available at
\url{https://github.com/leemgs/sage}. The repository README records exact
commands; \texttt{SHA256SUMS.txt} records packaged artefact checksums. A public
immutable release and archival DOI remain pre-acceptance requirements.

%% ---- Author Contributions ---------------------------------
\section*{Author Contributions}
G.L.: Conceptualization, Formal analysis, Investigation, Methodology,
Software, Validation, Visualization, Writing -- original draft,
Writing -- review \& editing.

%% ---- Competing Interests ----------------------------------
\section*{Competing Interests}
The author declares no competing interests.

%% ---- Acknowledgements -------------------------------------
\input{section/090_ack.tex}

%% ---- References -------------------------------------------
\input{section/095_reference_nature.tex}

%% ---- Supplementary Information --------------------------------
%% NMI: submitted as a SEPARATE file — see supplementary.tex
%% (builds supplementary.pdf; Supplementary Notes 1–5, Tables 1–11).

\end{document}
